Global patterns of forest autotrophic carbon fluxes

### 2020-08-03 manuscript/manuscript draft.Rmd
revisions to intro
- changed discussion around @huston_global_2009, which is pretty throughly refuted by @gillman_latitude_2015

Forest C fluxes decrease with latitude, but it remains unclear whether and how the shape of this relationship varies between fluxes. 
Fluxes are lowest in the boreal regions, and increase into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @piao_forest_2010; @jung_global_2011]. 
However, evidence is inconclusive on whether primary productivity continues to increase into the tropics--a question complicated by the fact that different studies use different measures of productivity to explore these relationships. 
Gross primary productivity ($GPP$) is highest in tropical forests [@beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies suggest that the highest values of net primary productivity ($NPP$) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], while others find $NPP$ highest in the tropics and decreasing with latitude [@simova_enigma_2017]. 
Other studies have found that aboveground net primary productivity ($ANPP$) decreases weakly with latitude [@huston_global_2009; @gillman_latitude_2015].

The latitudinal gradient in C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017]. The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. There is strong evidence that both MAT and MAP show significant positive relationships with C fluxes [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different C flux variables across studies. Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while $GPP$ increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases linearly with temperature. Taylor et al. [-@taylor_temperature_2017] showed a positive interaction between MAT and MAP in shaping tropical forest productivity, such that high rainfall had a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates. Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and carbon supply [@hursh_sensitivity_2017; @xu_contribution_2016]. There is evidence that C fluxes also respond to variables such as cloud cover [@taylor_temperature_2017], solar radiation [@fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005].

MAT and MAP are very coarse measures of climate, and so fail to capture much variation in climate on an intra-annual scale, including the effects of factors such as growing season length, number of frost-free days, temperature seasonality, and dry season length. Some studies have suggested that the apparently strong relationship between MAT and C fluxes is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @michaletz_convergence_2014; @michaletz_drivers_2018]. Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] find that, within the growing season, there is no significant relationship between net primary productivity (loosely defined to include a mix of **VARIABLES**) and MAT, suggesting that the influence of temperature may be limited to determining the length of the frost-free growing season.

The recent development of the Global Forest Carbon database (ForC), which synthesizes multiple variables and including records of stand history [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018], opens up the possibility for a standardized analysis of global scale variation in multiple C fluxes and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). First, we ask how nine forest autotropic carbon fluxes in ForC vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between C flux and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on C flux standardized by growing season length.

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3),"$^{,}$", footnote_marker_number(10))
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(9))
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4))
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(5))
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(6),"$^{,}$",footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(8))
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of research questions, corresponding hypotheses, and results. Statistically signficant support for/ rejection of hypotheses is indicated with 'yes'/'no', and '-' indicates no significant relationship. Parentheses indicate partial overall support or rejection of hypotheses across all fluxes considered.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do C fluxes vary with latitude?", 1, 1, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
  pack_rows("Q2. How do C fluxes vary with MAT and MAP?", 2, 4, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q3. How are C fluxes related to other climate variables?", 5, 7, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence C fluxes?", 8, 11, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do C fluxes vary with climate?", 12, 15, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Schuur (2003)", "Taylor et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Piao et al. (2010)", "Huston & Wolverton (2009)"), footnote_as_chunk = TRUE)
```

>>>>

Forest C fluxes decrease with latitude [e.g., @luyssaert_co_2007; @piao_forest_2010; @gillman_latitude_2015; @li_mapping_2019]; but it remains unclear whether and how the shape of this relationship varies among fluxes. 
For instance, gross primary productivity ($GPP$) increases continuously with decreasing latitude and is indisputably highest in tropical forests [@luyssaert_co_2007; @beer_terrestrial_2010; @jung_global_2011; @badgley_terrestrial_2019; @li_mapping_2019]. 
In contrast, some studies have suggested that net primary productivity ($NPP$), or its aboveground portion ($ANPP$), exhibits a less distinct increase from temperate to tropical forests [@luyssaert_co_2007]--or even a decrease [@huston_global_2009, but see @gillman_latitude_2015]. 
A shallower increase in $NPP$ than in $GPP$ with decreasing latitude would align with the suggestion that tropical forests tend to have low carbon use efficiency [$CUE$= $NPP$/$GPP$; @de_lucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016].
However, accurately comparing relationships of multiple forest C fluxes across latitude requires a large database with standardization for methodology, stand history, and management, which to date has not been possible.

The latitudinal gradient in forest C flux rates, along with altitudinal gradients [@girardin_net_2010; @malhi_variation_2017], is driven primarily by climate, which is a significant driver of C fluxes across broad spatial scales [@luyssaert_co_2007; @cleveland_relationships_2011; @hursh_sensitivity_2017].
The majority of studies have focused on exploring the relationships of C fluxes to mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. 
There is strong evidence that both MAT and MAP show significant positive relationships with C fluxes [@chu_does_2016]; however, as with latitude, we lack synthetic understanding of how these relationships vary across climate fluxes. 
C flux increases strongly with $MAT$ on the global scale, but whether they saturate or potentially decrease at higher temperatures remains disputed.
In the global data compilation of Luyssaert et al. [-@luyssaert_co_2007], $GPP$ increased linearly with MAT, whereas $NPP$ saturated at around 10$^\circ$C MAT. 
In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in $GPP$ saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] finds that $NPP$ increases continuously with temperature. 
Various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2,445 mm MAP [@schuur_productivity_2003]. 
Interactions between MAT and MAP are also possible; within the tropics, there is a positive interaction between MAT and MAP in shaping $ANPP$, such that high rainfall has a negative effect on productivity in cooler climates, compared to a positive effect in warmer climates [@taylor_temperature_2017]. 
Such complicated dynamics play out with belowground fluxes as well, with typically dominant soil temperature mediated by moisture and C supply [@hursh_sensitivity_2017; @xu_contribution_2016]. 
There is also evidence that C fluxes also respond to climate variables such as cloud cover [@taylor_temperature_2017], solar radiation [@beer_terrestrial_2010; @fyllas_solar_2017], and potential evapotranspiration [@kerkhoff_plant_2005]; however, these are not typically assessed in global-scale analyses of annual forest C flux.

As metrics of annual climate, MAT and MAP fail to capture variation in climate on an intra-annual scale, including temperature and precipitation seasonality and growing season length. 
Some studies have suggested that the apparently strong relationship between MAT and C fluxes is actually a consequence of the correlation between MAT and growing season length [@kerkhoff_plant_2005; @michaletz_convergence_2014; @michaletz_drivers_2018]. 
Kerkhoff et al. [-@kerkhoff_plant_2005] and Michaletz et al. [-@michaletz_convergence_2014] found no significant relationship between growing season temperature and net primary productivity (loosely defined to include a mix of **VARIABLES**) standardized to growing season length.
While this suggests that the influence of temperature may be limited to determining the length of the frost-free growing season, *analysis with clearly defined variables and standardized forest ages would be necessary to test the the veracity and generality of this hypothesis.

The recent development of the Global Forest Carbon database (ForC), which synthesizes multiple variables and including records of stand history [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018], opens up the possibility for a standardized analysis of global scale variation in multiple C fluxes and the principle climatic drivers of these patterns.
In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1).
First, we ask how nine forest autotrophic carbon fluxes in ForC vary with latitude.
We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables.
Finally, we consider the relationship between C flux and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on C flux standardized by growing season length.

\renewcommand{\arraystretch}{2}
```{r eval = TRUE, echo=FALSE, warning=FALSE}
library(knitr)
library(kableExtra)
hypothesis_table <- read.csv("hypothesis_table.csv", stringsAsFactors = FALSE, check.names = FALSE)
hypothesis_table[1,1] <- paste0(hypothesis_table[1,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(2),"$^{,}$", footnote_marker_number(3),"$^{,}$", footnote_marker_number(10))
hypothesis_table[2,1] <- paste0(hypothesis_table[2,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4),"$^{,}$",footnote_marker_number(9))
hypothesis_table[3,1] <- paste0(hypothesis_table[3,1], footnote_marker_number(1),"$^{,}$",footnote_marker_number(4))
hypothesis_table[4,1] <- paste0(hypothesis_table[4,1], footnote_marker_number(5))
hypothesis_table[10,1] <- paste0(hypothesis_table[10,1], footnote_marker_number(6),"$^{,}$",footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[11,1] <- paste0(hypothesis_table[11,1], footnote_marker_number(7),"$^{,}$", footnote_marker_number(8))
hypothesis_table[12,1] <- paste0(hypothesis_table[12,1], footnote_marker_number(8))
kable(hypothesis_table, format = "latex", booktabs = TRUE, caption = "Summary of research questions, corresponding hypotheses, and results. Statistically signficant support for/ rejection of hypotheses is indicated with 'yes'/'no', and '-' indicates no significant relationship. Parentheses indicate partial overall support or rejection of hypotheses across all fluxes considered.", escape = F) %>%
  kable_styling(latex_options = c("scale_down", "hold_position"), font_size = 12) %>%
  pack_rows("Q1. How do C fluxes vary with latitude?", 1, 1, latex_gap_space = "1em", colnum = 5, hline_before = FALSE) %>%
  pack_rows("Q2. How do C fluxes vary with mean annual temperature (MAT) and precipitation (MAP)?", 2, 4, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q3. How are C fluxes related to other annual climate variables?", 5, 7, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q4. How does seasonality influence annual C fluxes?", 8, 11, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  pack_rows("Q5. When standardised by growing season length, how do annual C fluxes vary with climate?", 12, 15, latex_gap_space = "1em", colnum = 5, hline_before = TRUE) %>%
  add_header_above(c(" ",  " ", "Forest autotrophic carbon fluxes" = 9, " ")) %>%
  #column_spec(2:11, width = "1.8cm") %>%
  kableExtra::landscape() %>%
  kableExtra::footnote(number = c("Luyssaert et al. (2007)", "Gillman et al. (2015)", "Simova and Storch (2017)", "Schuur (2003)", "Taylor et al. (2016)", "Malhi (2012)", "Michaletz et al. (2014)", "Chu et al. (2016)", "Piao et al. (2010)", "Huston & Wolverton (2009)"), footnote_as_chunk = TRUE)
```
